Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-05-03 paperDistance/paper.tex
paper tweaks

\author{Jo Bovy}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
\affil{Department of Astronomy and Astrophysics, University of Toronto, 50 St. George Street, Toronto, ON M5S 3H4, Canada}

\author{Boris Leistedt}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}

\author{Adrian Price-Whelan}
\affil{Princeton University Observatory, Princeton University, 4 Ivy Lane, Princeton, NJ 08544, USA}

@@

Although we have not performed principled hierarchical Bayesian inference in this
project, we built our prior for each star's distance by
performing a statistically responsible deconvolution
of the \gaia\ \tgas\ data that is justifiable under a clear set of
assumptions.
This deconvolution makes use of a reasonable noise model and an
assumption of stationarity to shrink the parallax uncertainties, and
do so dramatically for the stars measured at lower signal-to-noise.
Because the method we use (Extreme Deconvolution or \xd, which is
an Empirical Bayesian maximum-marginalized-likelihood estimator) accounts for
heteroskedastic noise, we were able to build our photometric parallax
model---which is a model of the \cmd---using
all the data, not just the data with the highest signal-to-noise or
largest parallaxes.
That is, the model for the \cmd\ we have built is representative for
the \tgas\ Catalog.
Inasmuch as that is true, we expect any (sensible) distance estimates
we generate from our distance posterior pdfs to be weakly biased.

The Empirical Bayesian methodology is a good approximation to full
hierarchical inference in the limit of large numbers of data points.
That is, when no individual star is carrying a lot of weight in the
model of the \cmd.
With its millions of stars, the \tgas\ Catalog may appear to safely
live in this limit.
However, because there are parts of the \cmd\ that are poorly populated,
even in \tgas, the Empirical Bayes approximation may be bad for some
color ranges.
In particular, there appear to be essentially no white dwarf stars,
and other classes (like hot main-sequence stars) have few to no exemplars
in the data set at high signal-to-noise (in parallax).
For all these reasons, we do think that this approximation is not perfectly
safe, and it is a goal to explore full hierarchical modeling in the coming
years (and see, for example, Leistedt et al. in preparation).

@@

The outputs of this project include posterior pdfs and posterior
samplings for stellar parallaxes or distances.
It is tempting to treat these outputs as equivalent to some kind of catalog of
measurements, as we treated the original parallax measurements from \tgas.
That is, it is tempting to treat these as simply ``better'' measurements.
Taken one star at a time, this is permitted and true, and the basis for our
claim that we have de-noised the \tgas\ Catalog.
However, there are important differences, as there are with all
probabilistic catalogs (\citealt{hogg11}; \citealt{portillo17}), between the
measurements and the de-noised measurements.
The most important difference, for users of our output, between our
output and the \tgas\ data input, is that the \tgas\ data is the representation
of a likelihood function, whereas our ouptut is a
representation of a posterior pdf, one per star.
If used naively, posterior information can lead to serious statistical
errors, because each data point has had a prior multiplied in.
Effectively multiplicative uses of the data will effectively take the prior
pdf to a large power.
We have (in previous work)
given examples of correct uses of posterior samplings or pdfs
for subsequent inferences (\citealt{hogg08}; \citealt{dfm14}); we
encourage power-users to consult those methodological
contributions before engaging with these outputs.

@@

There are other prior information options when inferring distances to stars. A commonly used prior for inferring distances (cite: Tri, CBJ) is the exponentially declining spatial density of stars. It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so the posteriors are also very smooth. It is a fairly weak prior and basically sets a smooth scale at which it's challenging for stars to lie at, and also allows for negative parallaxes. It is a fairly broad prior so the median \gaia\ star has an increased posterior variance. It has the largest affect on the very noisy and neegative parallaxes and has the nice property of bringing these noisy or negative measurements to reasonable distances. Given this prior, what does the observed \cmd\ look like?  The \cmd\ is broad and the red clump highly stretched.

@@

Come back to the assumptions. What happens if we change or violate them? Long discussion of the noise model. We took serious short-cuts. Call out to Boris.

LMA: Some discussion of the inconsistency of how we treat dust within the probabilistic framework.
=======
Come back to the assumptions. What happens if we change or violate them?

@@

\subsection{Proper Dust Corrections}
Within our probabilistic framework, the way we handle dust is not proper. We are taking a point estimate of the probabilistic dust model at a quantile distance of our posterior, correcting our photometry using this point estimate and rederiving the prior in \xd. A full, proper account would sample the dust model at distances sampled from the posterior and propagate those samples back into \xd during the iterative proccess. The issues is that this dust sample distribution would not be a Gaussian distribution which \xd\ requires. We could calculate this more complex dust distribution and estimate it as a Gaussian distribution to feed to \xd, however, this is beyond the scope of this paper.

@@

\acknowledgments It is a pleasure to thank
  Ana Bonaca (Harvard),
  Andy Casey (Monash),
  Dustin Lang (Toronto) for cross matched 2MASS-TGAS catalogue
and the attendees at the Stars Group Meeting at the Flatiron Institute
Center for Computational Astrophysics for comments and input.
This project was developed in part at the 2016 \acronym{NYC} Gaia Sprint, hosted
by the Center for Computational Astrophysics at the Simons Foundation
in New York City.

>>>>

%\author{Jo Bovy}
%\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
%\affil{Department of Astronomy and Astrophysics, University of Toronto, 50 St. George Street, Toronto, ON M5S 3H4, Canada}

%\author{Boris Leistedt}
%\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}

%\author{Adrian Price-Whelan}
%\affil{Princeton University Observatory, Princeton University, 4 Ivy Lane, Princeton, NJ 08544, USA}

@@

Although we have not performed principled hierarchical Bayesian inference in this
project, we built our prior for each star's distance by
performing a statistically responsible deconvolution
of the \gaia\ \tgas\ data that is justifiable under a clear set of
assumptions.
This deconvolution makes use of a reasonable noise model and an
assumption of stationarity to shrink the parallax uncertainties, and
do so dramatically for the stars measured at lower signal-to-noise.
Because the method we use (Extreme Deconvolution or \xd, which is
an Empirical Bayesian maximum-marginalized-likelihood estimator) accounts for
heteroskedastic noise, we were able to build our photometric parallax
model---which is a model of the \cmd---using
all the data, not just the data with the highest signal-to-noise or
largest parallaxes.
That is, the model for the \cmd\ we have built is representative for
(a subset of) the \tgas\ Catalog.
Inasmuch as that is true, we expect any (sensible) distance estimates
we generate from our distance posterior pdfs to be only weakly biased.

The Empirical Bayesian methodology is a good approximation to full
hierarchical inference in the limit of large numbers of data points.
That is, when no individual star is carrying a lot of weight in the
model of the \cmd.
With its millions of stars, the \tgas\ Catalog may appear to safely
live in this limit.
However, because there are parts of the \cmd\ that are poorly populated,
even in \tgas, the Empirical Bayes approximation may be bad for some
color ranges.
In particular, there appear to be essentially no white dwarf stars,
and other classes (like bright red giant branch stars) have few to no exemplars
in the data set at high signal-to-noise (in parallax).
For all these reasons, we do think that this approximation is not perfectly
safe, and it is a goal to explore full hierarchical modeling in the coming
years (and see, for example, Leistedt et al. in preparation).

@@

The outputs of this project include posterior pdfs and posterior
samplings for stellar parallaxes or distances.
It is tempting to treat these outputs as equivalent to some kind of catalog of
measurements, as we treated the original parallax measurements from \tgas.
That is, it is tempting to treat these as simply ``better'' measurements.
Taken one star at a time, this is permitted and true, and the basis for our
claim that we have de-noised the \tgas\ Catalog.
However, there are important differences, as there are with all
probabilistic catalogs (\citealt{hogg11}; \citealt{portillo17}), between the
measurements and the de-noised measurements.
The most important difference, for users of our output, between our
output and the \tgas\ data input, is that the \tgas\ data is the representation
of a likelihood function, whereas our ouptut is a
representation of a posterior pdf, one per star.
If used naively, posterior information can lead to serious statistical
errors, because each data point has had a prior multiplied in.
Multiplicative uses of the data will effectively take the prior
pdf to a large power.
We have (in previous work)
given examples of correct uses of posterior samplings or pdfs
for subsequent inferences (\citealt{hogg08}; \citealt{dfm14}); we
encourage power-users to consult those methodological
contributions before engaging with these outputs.

@@

There are other prior information options when inferring distances to stars. One option is the exponentially declining spatial density of stars (cite: Tri, CBJ). It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so the posteriors are also very smooth. It is a fairly weak prior and basically sets a smooth scale at which it's challenging for stars to lie at, and also allows for negative parallaxes. It is a fairly broad prior so the median \gaia\ star has an increased posterior variance. It has the largest affect on the most noisy and negative parallaxes and has the nice property of bringing these noisy or negative measurements to reasonable parallaxes. Given this prior, what does the observed \cmd\ look like?  The \cmd\ is broad and the red clump highly stretched.

@@

Come back to the assumptions. What happens if we change or violate them?

@@

\subsection{Proper Dust Corrections}
Within our probabilistic framework, we improperly handle dust. We are taking a point estimate of the probabilistic dust model at a quantile distance of our posterior, correcting our photometry using this point estimate and rederiving the prior in \xd. A full, proper account would sample the dust model at distances sampled from the posterior and propagate those samples back into \xd during the iterative proccess. The issues is that this dust sample distribution would not be a Gaussian distribution which \xd\ requires. We could calculate this more complex dust distribution and estimate it as a Gaussian distribution to feed to \xd, however, this is beyond the scope of this paper.

@@

\acknowledgments It is a pleasure to thank
  Ana Bonaca (Harvard),
  Andy Casey (Monash),
  Dustin Lang (Toronto) for cross matched 2MASS-TGAS catalogue
  Jo Bovy (Toronto)
  Boris Leistedt (NYU)
  Adrian Price-Whelan (Princeton)
and the attendees at the Stars Group Meeting at the Flatiron Institute
Center for Computational Astrophysics for comments and input.
This project was developed in part at the 2016 \acronym{NYC} Gaia Sprint, hosted
by the Center for Computational Astrophysics at the Simons Foundation
in New York City.
